Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North

### 2024-01-29 reports/GnC_report/paper/paper.qmd
update docs website

---
title: "Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North"
subtitle: "Prepared for: Canadian Wildlife Service, Canadian Forest Service, and Wilfrid Laurier University"
author:
  - name: Valentin Lucet
    email: valentin.lucet@gmail.com
    orcid: 0000-0003-0268-818X
    affiliation: 
      - ref: wlu
  - name: Samantha McFarlane
    affiliation:
      - ref: eccc
  - name: Jennifer Baltzer
    affiliation:
      - ref: wlu
  - name: Cheryl A. Johnson
    affiliation:
      - ref: eccc
  - name: Mathieu Leblond
    affiliation:
      - ref: eccc
  - name: Eric Neilson
    affiliation:
      - ref: nrcan
  - name: Philip Weibe
    affiliation:
      - ref: nrcan
  - name: Frances Stewart
    orcid: 0000-0001-9344-8346
    affiliation:
      - ref: wlu
  - name: Josie Hughes
    affiliation:
      - ref: eccc
affiliations:
  - id: wlu
    name: Wilfrid Laurier University
  - id: eccc
    name: Environment & Climate Change Canada
  - id: nrcan
    name: Natural Resources Canada
title-block-published: "Last updated"  
date: now
date-format: long
format:
  html: default
  # pdf:
  #   toc: true
  #   number-sections: true
  #   colorlinks: true
  #   extra_dependencies: ["float"]
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
bibliography: references.bib
abstract: >
  Northern Ontario holds some of the world’s last intact wild places, but it is under pressure from resource extraction and climate change. Our understanding of biodiversity measures like species occurrence, abundance, diversity, and density for the region are lacking, yet needed for strategic decision-making about the impacts of resource extraction on ecosystem services and biodiversity. The use of non-invasive survey methods, such as camera trap arrays, has significantly changed our ability to quantify wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification of species. Machine learning methods, such as deep neural networks, are a promising tool to overcome this data bottle-neck. Yet, variables like strong seasonality, changes in vegetation type, and species characteristics can strongly impact their efficacy. The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed across an heterogeneous sampling area straddling two Northern Ontario ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We’ve recently developed an open-source workflow for camera trap image identification, access, and storage. We continue to use this workflow for ongoing image identification while investigating requirements to develop a machine learning image classifier.

@@

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '~/Documents/WILDLab/rofcamtrap/')
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
backup_options <- options()
options(scipen = 1, digits = 2)
```

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored.

With the following code you can access and display values from the yml header above.
Keywords: `r rmarkdown::metadata$keywords`
Highlights: `r rmarkdown::metadata$highlights`

Here is a citation [@Marwick2017] -->

@@

```{r}
#| label: spatial-basics
#| echo: false

@@

# zones <- 
#   sf::st_read("analysis/data/raw_data/ecozones/ECOREGN/LIO-2022-10-25/ECOREGION.shp",
#               quiet=TRUE, ) |> 
#   sf::st_transform(crs = sf::st_crs(cams_sf))

@@

# regions_labels <- zones_filter |> 
#   sf::st_drop_geometry() |> 
# dplyr::bind_cols(data.frame(x = c(-88.5, -84, -88.5),
#                             y = c(55, 52, 52.6)))

regions_labels <- zones_filter |> 
  sf::st_drop_geometry() |> 
  dplyr::bind_cols(data.frame(x = c(-84.5, -88.5),
                              y = c(52, 52.6)))

map <- ggmap::get_stadiamap(bbox = c(left = -90, bottom = 50, right = -83, top = 55.5), 
                            zoom = 9, maptype = "stamen_terrain")
```

```{r}
#| label: fig-deployment
#| echo: false
#| fig-cap: "Camera deployment status for the 193 cameras, 170 of which were retrieved in September 2022. Labels for the 3 ecoregions spanned by the sampling area are included."

base <- ggmap::ggmap(map) +
  ggplot2::labs(x = "Longitude", y = "Latitude") +
  # ggplot2::geom_polygon(data = pols, ggplot2::aes(x=X, y=Y), col = 1, fill = NA)
  ggplot2::geom_path(data = pols, ggplot2::aes(x = X, y = Y), col = 1, linewidth = 0.2)

p <- base +
  # ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOREGION),inherit.aes = FALSE,
  #                     size = 2.5, data = regions_labels) +
  ggplot2::geom_label(ggplot2::aes(x = x, y = y, label = ECOZONE_NA), inherit.aes = FALSE,
                      size = 2.5, data = regions_labels) +
  ggplot2::geom_point(data = cams_sf_coords, #cex = 1, 
                      shape = 24, col = "black", fill = "red",
                      ggplot2::aes(x = X, y = Y, #fill=is_retrieved
                                   ), 
                      inherit.aes = FALSE) # +
  # ggplot2::labs(fill = "Camera \nRetrieved")
p
ggplot2::ggsave(plot = p, filename = "figures/camera_array.png", 
                width = 6, height = 8, dpi = 300)
ggplot2::ggsave(plot = p, filename = "reports/presentations/rofcamtrap-nov2023/images/camera_array.png", 
                width = 6, height = 8, dpi = 300)
```

```{r}
#| label: fig-deployment-poster
#| eval: false
#| echo: false

@@

```{r}
#| label: load-data
#| echo: false

anns_wide <- readRDS("data/objects/annotations_wide_treated.rds")
dets <- readRDS("data/objects/detections.rds")

@@

# Hardcoded check
stopifnot(nb_image == 3288178) # 2788375

exif <- readr::read_csv("data/tabular/all_exif.csv") |>
  janitor::clean_names() |>
  dplyr::mutate(image_id = stringr::str_replace_all(source_file, "/", "_")) |>
  dplyr::mutate(image_id = stringr::str_replace_all(image_id, ".JPG", "")) |>

    dplyr::mutate(source_file = stringr::str_remove(source_file, "/media/vlucet/TrailCamST/TrailCamStorage2/")) |>
  dplyr::mutate(source_file = stringr::str_remove(source_file, "/media/vlucet/TrailCamST/TrailCamStorage/")) |>

  dplyr::mutate(source_file = stringr::str_remove(source_file, 
                                               "/workspace/storage/TrailCamStorage_2/")) |>
  dplyr::mutate(source_file = stringr::str_remove(source_file, 
                                               "/workspace/storage/TrailCamStorage/")) |>

  dplyr::mutate(image_id = stringr::str_remove(image_id, "_media_vlucet_TrailCamST_TrailCamStorage2_")) |>
  dplyr::mutate(image_id = stringr::str_remove(image_id, "_media_vlucet_TrailCamST_TrailCamStorage_")) |>

  dplyr::mutate(image_id = stringr::str_remove(image_id, 
                                               "_workspace_storage_TrailCamStorage_2_")) |>
  dplyr::mutate(image_id = stringr::str_remove(image_id, 
                                               "_workspace_storage_TrailCamStorage_")) |>

  dplyr::mutate(image_id = strsplit(image_id, "_") |> 
                  purrr::map(~paste0(.x[2:length(.x)], collapse = "_")) |> 
                  unlist()) |>

  dplyr::mutate(
    source_file = 
      stringr::str_remove(source_file, 
                          stringr::fixed("/media/vlucet/TrailCamST/TrailCamStorage/"))) |> 
  dplyr::mutate(
    source_file = 
      stringr::str_remove(source_file, 
                          stringr::fixed("/media/vlucet/TrailCamST/TrailCamStorage_2/")))
  # dplyr::mutate(image_id = stringr::str_sub(image_id, category6))
```

Images were retrieved from all cameras using a 32 GB SD card and copied onto a 10 TB hard drive (LaCie d2 Professional). The images amounted to 1.8TB of memory on this disk, for a total of __`r nb_image_fmt`__ images. Images were stored with the following directory structure:

@@

```{r}
#| label: basics-stats-1
#| echo: false

@@

At this first step into the process, no filtering was applied beyond the model predictions which classifies images as either likely to contain an object of relevance or empty. About __`r dets_tb_percent[1]` %__ of the images, (__`r dets_tb_fmt[1]`__) images, were found by the model to possibly contain either a person, animal, or vehicle. The model did not produce detections for __`r dets_tb_fmt[2]`__ images, (__`r dets_tb_percent[2]` %__) of the total.

```{r}
#| label: basics-stats-2
#| echo: false
dets_0 <- dets |> dplyr::filter(confidence > 0)
dets_0.1 <- dets |> dplyr::filter(confidence > 0.1)

@@

Because the model can produce more than one detection per image, this resulted in a total of __`r dets_tb_cat_unfil_sum_fmt`__ detections. The totals for each category are: __`r dets_tb_cat_unfil_fmt[1]`__ animals (or __`r dets_tb_cat_unfil_percent[1]` %__); __`r dets_tb_cat_unfil_fmt[2]`__ persons (or __`r dets_tb_cat_unfil_percent[2]` %__); and __`r dets_tb_cat_unfil_fmt[3]`__ vehicles (or __`r dets_tb_cat_unfil_percent[3]` %__).

@@

Using a cut-off value for the detection confidence of 0.1 (,i.e. discarding all detections with confidence below 0.1), this left us with a total of __`r dets_tb_cat_sum_fmt`__ detections to sort through (see @fig-detections-histogram for an histogram of the detection scores, grouped by detection types). After applying this confidence threshold (decided on by the Wildlife Working Group) the total images to process  for each category are:  __`r dets_tb_cat_fmt[1]`__ animals (__`r dets_tb_cat_percent[1]` %__); __`r dets_tb_cat_fmt[2]`__ persons (__`r dets_tb_cat_percent[2]` %__); and __`r dets_tb_cat_fmt[3]`__ vehicles (__`r dets_tb_cat_percent[3]` %__).

```{r}
#| label: fig-detections-histogram
#| echo: false
#| fig-cap: "Histogram of detection confidence scores produced for each bounding box predicted by Megadetector, showing a bimodal distribution. The colors represent the three categories of object identified by Megadetector: animals, persons and vehicles. Note that the hsitogram is truncated at our selected confidence threshold of 0.1."
#| fig.width: 8
#| fig.height: 4.5

@@

```{r}
#| label: basics-stats-3
#| echo: false

@@

# Filter for non empty images
anns_wide_noempty <- anns_wide |>
  dplyr::filter(!is.na(to_name))

saveRDS(anns_wide_noempty, "data/objects/annotations_wide_noempty.rds")

@@

Out of the __`r dets_tb_cat_fmt[1]`__ animal detections, __`r processed_fmt`__ have been processed to date using our platform (or __`r processed_pct` %__) for a total of __`r processed_img_fmt`__ images (i.e. __`r processed_img_pct` %__ of the total amount of images, or __`r processed_img_up_pct` %__ of the total amount of images uploaded to the platform).

As stated above, all images of a given burst are viewed if at least one image in the burst contains a detections with confidence above the threshold. This include potentially empty images. If we add the empty images viewed as part of a burst, it brings the total amount to __`r viewed_fmt`__ images, (i.e. __`r viewed_pct` %__ of the total amount of images, or __`r viewed_pct_up` %__ of the total amount of images uploaded to the platform).

```{r}
#| label: mans
#| echo: false

@@

Out of the __`r processed_fmt`__ detections, __`r dels_nb_fmt`__ were false positives (or __`r dels_nb_pct` %__), where Megadetector identified something, but there was nothing. In addition, __`r non_mans_nb_fmt`__ were true positives (or __`r non_mans_nb_pct` %__), to which __`r mans_nb_fmt`__ false negatives were added (manually added bounding boxes, __`r mans_nb_pct` %__). Out of the __`r non_mans_nb_fmt`__ true positives, __`r orig_tb_fmt[3]`__ had to have their bounding box adjusted. The table below summarises information for the tagging progress.

```{r}
#| label: tab-summary-table
#| echo: false

@@

We are committed to producing a reproducible workflow. All code necessary to reproduce this report is available at an online [repository](https://github.com/StewartWILDlab/rofcamtrap). The two software tools used in our workflow are also available at online repositories:

- [**mdtools**](https://github.com/StewartWILDlab/mdtools): a python-based command line tool for converting Megadetector outputs into LabelStudio inputs, and LabelStudio outputs into CSV tables.

- [**mdscripts**](https://github.com/StewartWILDlab/mdscripts): a bash-based command line tool for running Megadetector and mdtools.

@@

```{r}
#| label: quality-control
#| echo: false

# Only annotations for animals
# TODO !!!! investigate weird id of cut length 1
anns_wide_animals <- anns_wide_noempty |>
  dplyr::filter(species != 'Human') |> 
  dplyr::filter(species != 'Vehicle') |> 
  dplyr::mutate(id_cut_length = purrr::map_dbl(id , \(x) {
    x |> stringr::str_split("_") |> purrr::map_dbl(~length(.x))
  })) |> 
  dplyr::filter(id_cut_length > 1) |> 
  dplyr::mutate(camera_id = purrr::map_chr(id , \(x) {
    x |>
      stringr::str_split("_") |>
      purrr::map(~.x[[2]]) |>
      unlist()
  }))

@@

anns_wide_animals_joined <- anns_wide_animals |>
  dplyr::left_join(exif, by = c("source_file", "image_id"))

@@

# anns_wide_animals_joined_dets <- anns_wide_animals |> 
#   dplyr::right_join(dets_images, by = "image_id")
```

@@

```{r eval=FALSE}
# library(forcats)
library(ggplot2)
# anns_wide_animals_joined_mut |> 
#   ggplot(aes(x=date_time, y=species)) + geom_point()
# anns_wide_animals_joined_mut |> 
#   dplyr::filter(species == "Fisher") |> View()

@@

```{r eval=FALSE}
# for_vic <- anns_wide_animals_joined_mut |> 
#   dplyr::select(id,species,date_time, camera_id) |> 
#   dplyr::left_join(cams_sf |> dplyr::select(camera_id, plot_id) |> sf::st_drop_geometry(),
#                    by = "camera_id")
# tot_camera <- for_vic$camera_id |> unique() |> length()
# hit_rates <- for_vic |> dplyr::select(species, camera_id) |> 
#   dplyr::group_by(species) |> dplyr::summarise(n_cameras=length(unique(camera_id))) |> 
#   dplyr::ungroup() |> dplyr::mutate(rate = n_cameras/tot_camera*100) |> 
#   dplyr::arrange(dplyr::desc(rate))
# # det_history_two_weeks <- 
# 
# readr::write_csv(for_vic, "../data/derived_data/observations_by_camera_2023Aug15.csv")
# readr::write_csv(hit_rates, "../data/derived_data/hit_rates_by_species_2023Aug15.csv")
```

```{r}
#| label: species
#| echo: false
anns_wide_animals_table <- anns_wide_animals |>
  dplyr::select(camera_id, species) |> 
  dplyr::filter(species != "Unknown") |>
  table() 
anns_wide_animals_table <- anns_wide_animals_table > 0
anns_wide_animals_table <- anns_wide_animals_table + 0
anns_wide_animals_table <- anns_wide_animals_table |> 
  as.data.frame() |> 
  tibble::rownames_to_column("camera_id") |> 
  dplyr::arrange(camera_id)
anns_wide_animals_table_tb <- anns_wide_animals_table |> 
  dplyr::select(-camera_id)
rownames(anns_wide_animals_table_tb) <- anns_wide_animals_table$camera_id
```

@@

base +
  ggplot2::coord_equal(xlim = c(-88.5, -84),
                       ylim = c(50, 54)) +
  scatterpie::geom_scatterpie(data = cams_sf_with_props,
                              ggplot2::aes(x=X, y=Y, group=plot_id),
                              cols=names(cams_sf_with_props)[4:(4+tmp_col-1)],
                              sorted_by_radius = TRUE,
                              legend_name = "Species",
                              pie_scale = 1.5) +
  ggplot2::labs(x="Lat", y="Long") +
    ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOZONE_NA),inherit.aes = FALSE,
                      size = 2.5, data = regions_labels_bis)
```

@@

```{r, eval = FALSE}
#| label: heatmap
#| echo: false
#| eval: false

@@

# rast <- raster::rasterFromXYZ(preds)
# raster::plot(rast)
# rast[rast<0] <- NA
# raster::plot(rast)

base_cut +
  ggplot2::geom_density2d_filled(
    data = cams_sf_sp_filt, ggplot2::aes(X, Y),
    binwidth = 0.05, alpha = 0.5)

@@

```{r}
#| label: bubbleplots
#| echo: false

@@

@@

  gg <- base_cut +
    ggplot2::geom_point(data = dat, col="black",
                        ggplot2::aes(x=X, y=Y, size = count), 
                        inherit.aes = FALSE) +
    ggplot2::facet_wrap(~month) +
    ggplot2::labs(size="Days \ndetected",
                  title=sp)

  ggplot2::ggsave(gg, filename = paste0("analysis/figures/",sp,"_bplot.png"))
}
```

@@

![](../figures/Caribou_bplot.png)
\vspace{1pt}

![](../figures/Sandhill crane_bplot.png)
\vspace{1pt}

![](../figures/Moose_bplot.png)
\vspace{1pt}

![](../figures/Wolf_bplot.png)
\vspace{1pt}

>>>>

---
title: "Using Camera Traps and Computer Vision to Quantify Wildlife Diversity and Co-Occurrence across \nOntario’s Far North"
subtitle: "Prepared for: Canadian Wildlife Service, Canadian Forest Service, and Wilfrid Laurier University"
author:
  - name: Valentin Lucet
    email: valentin.lucet@gmail.com
    orcid: 0000-0003-0268-818X
    affiliation: 
      - ref: wlu
  - name: Samantha McFarlane
    affiliation:
      - ref: eccc
  - name: Jennifer Baltzer
    affiliation:
      - ref: wlu
  - name: Cheryl A. Johnson
    affiliation:
      - ref: eccc
  - name: Mathieu Leblond
    affiliation:
      - ref: eccc
  - name: Eric Neilson
    affiliation:
      - ref: nrcan
  - name: Philip Weibe
    affiliation:
      - ref: nrcan
  - name: Frances Stewart
    orcid: 0000-0001-9344-8346
    affiliation:
      - ref: wlu
  - name: Josie Hughes
    affiliation:
      - ref: eccc
affiliations:
  - id: wlu
    name: Wilfrid Laurier University
  - id: eccc
    name: Environment & Climate Change Canada
  - id: nrcan
    name: Natural Resources Canada
title-block-published: "Last updated"  
date: now
date-format: long
format:
  html: default
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
    extra_dependencies: ["float"]
format-links: false
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
bibliography: references.bib
abstract: >
  Northern Ontario holds some of the world’s last intact wild places, but it is under pressure from resource extraction and climate change. Our understanding of biodiversity measures like species occurrence, abundance, diversity, and density for the region are lacking, yet needed for strategic decision-making about the impacts of resource extraction on ecosystem services and biodiversity. The use of non-invasive survey methods, such as camera trap arrays, has significantly changed our ability to quantify wildlife across large and remote regions. However, these methods generate a large amount of data and require tools to speed up the identification of species. Machine learning methods, such as deep neural networks, are a promising tool to overcome this data bottle-neck. Yet, variables like strong seasonality, changes in vegetation type, and species characteristics can strongly impact their efficacy. The extent to which these methods perform for northern regions is not well established. Using an array of cameras deployed across an heterogeneous sampling area straddling two Northern Ontario ecozones (Hudson Bay Lowlands and Ontario Shield), we first look to quantify current species diversity and co-occurrence. We’ve recently developed an open-source workflow for camera trap image identification, access, and storage. We continue to use this workflow for ongoing image identification while investigating requirements to develop a machine learning image classifier.

@@

```{r, setup, include=FALSE}
root_dir <- here::here()
paper_dir <- here::here("reports", "GnC_report", "paper")
knitr::opts_knit$set(root.dir = root_dir)
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")
backup_options <- options()
options(scipen = 1, digits = 2)
do_eval <- FALSE
```

@@

```{r echo=FALSE}
#| label: spatial-basics

@@

@@

regions_labels <- zones_filter |> 
  sf::st_drop_geometry() |> 
  dplyr::bind_cols(data.frame(x = c(-84.5, -88.5),
                              y = c(52, 52.6)))
```

```{r eval=do_eval, echo=FALSE}
#| label: spatial-compute

map <- ggmap::get_stadiamap(bbox = c(left = -90, bottom = 50, right = -83, top = 55.5), 
                            zoom = 5, maptype = "stamen_terrain")

base <- ggmap::ggmap(map) +
  ggplot2::labs(x = "Longitude", y = "Latitude") +
  ggplot2::geom_path(data = pols, ggplot2::aes(x = X, y = Y), col = 1, linewidth = 0.2)
saveRDS(base, here::here(paper_dir, "base_map.rds"))

p <- base +
  ggplot2::geom_label(ggplot2::aes(x = x, y = y, label = ECOZONE_NA), inherit.aes = FALSE,
                      size = 2.5, data = regions_labels) +
  ggplot2::geom_point(data = cams_sf_coords,
                      shape = 24, col = "black", fill = "red",
                      ggplot2::aes(x = X, y = Y), 
                      inherit.aes = FALSE)
ggplot2::ggsave(plot = p, filename = "figures/camera_array.png", 
                width = 6, height = 8, dpi = 300)
saveRDS(p, here::here(paper_dir, "camera_array.rds"))
```

```{r echo=FALSE}
#| label: fig-deployment
#| fig-cap: "Camera deployment status for the 193 cameras, 170 of which were retrieved in September 2022. Labels for the 3 ecoregions spanned by the sampling area are included."
p <- readRDS(here::here(paper_dir, "camera_array.rds"))
p
```

```{r eval=do_eval, echo=FALSE}
#| label: fig-deployment-poster

@@

```{r eval=do_eval, echo=FALSE}
#| label: process-exif

exif <- readr::read_csv("data/tabular/all_exif.csv") |>
  janitor::clean_names() |>
  dplyr::mutate(image_id = stringr::str_replace_all(source_file, "/", "_")) |>
  dplyr::mutate(image_id = stringr::str_replace_all(image_id, ".JPG", "")) |> 
  dplyr::mutate(source_file = stringr::str_remove(source_file, "/media/vlucet/TrailCamST/TrailCamStorage2/")) |>
  dplyr::mutate(source_file = stringr::str_remove(source_file, "/media/vlucet/TrailCamST/TrailCamStorage/")) |> 
  dplyr::mutate(source_file = stringr::str_remove(source_file, "/workspace/storage/TrailCamStorage_2/")) |>
  dplyr::mutate(source_file = stringr::str_remove(source_file, "/workspace/storage/TrailCamStorage/")) |> 
  dplyr::mutate(image_id = stringr::str_remove(image_id, "_media_vlucet_TrailCamST_TrailCamStorage2_")) |>
  dplyr::mutate(image_id = stringr::str_remove(image_id, "_media_vlucet_TrailCamST_TrailCamStorage_")) |> 
  dplyr::mutate(image_id = stringr::str_remove(image_id, "_workspace_storage_TrailCamStorage_2_")) |>
  dplyr::mutate(image_id = stringr::str_remove(image_id, "_workspace_storage_TrailCamStorage_")) |> 
  dplyr::mutate(image_id = strsplit(image_id, "_") |> 
                  purrr::map(~paste0(.x[2:length(.x)], collapse = "_")) |> 
                  unlist()) |> 
  dplyr::mutate(source_file = stringr::str_remove(source_file, stringr::fixed("/media/vlucet/TrailCamST/TrailCamStorage/"))) |> 
  dplyr::mutate(source_file = stringr::str_remove(source_file, stringr::fixed("/media/vlucet/TrailCamST/TrailCamStorage_2/")))

readr::write_csv(exif, here::here("data", "tabular", "exif_processed.csv"))
```

```{r echo=FALSE}
#| label: load-data

anns_wide <- readRDS("data/objects/annotations_wide_treated.rds")
anns_wide_noempty  <- readRDS("data/objects/annotations_wide_noempty.rds")
dets <- readRDS("data/objects/detections.rds")
exif <- readr::read_csv(here::here("data", "tabular", "exif_processed.csv"))

@@

# Hardcoded check
stopifnot(nb_image == 3288178) # 2788375
```

Images were retrieved from all cameras using a 32 GB SD card and copied onto a 10 TB hard drive (LaCie d2 Professional). The images amounted to 1.8TB of memory on this disk, for a total of **`r nb_image_fmt`** images. Images were stored with the following directory structure:

@@

```{r echo=FALSE}
#| label: basics-stats-1

@@

At this first step into the process, no filtering was applied beyond the model predictions which classifies images as either likely to contain an object of relevance or empty. About **`r dets_tb_percent[1]` %** of the images, (**`r dets_tb_fmt[1]`**) images, were found by the model to possibly contain either a person, animal, or vehicle. The model did not produce detections for **`r dets_tb_fmt[2]`** images, (**`r dets_tb_percent[2]` %**) of the total.

```{r echo=FALSE}
#| label: basics-stats-2

dets_0 <- dets |> dplyr::filter(confidence > 0)
dets_0.1 <- dets |> dplyr::filter(confidence > 0.1)

@@

Because the model can produce more than one detection per image, this resulted in a total of **`r dets_tb_cat_unfil_sum_fmt`** detections. The totals for each category are: **`r dets_tb_cat_unfil_fmt[1]`** animals (or **`r dets_tb_cat_unfil_percent[1]` %**); **`r dets_tb_cat_unfil_fmt[2]`** persons (or **`r dets_tb_cat_unfil_percent[2]` %**); and **`r dets_tb_cat_unfil_fmt[3]`** vehicles (or **`r dets_tb_cat_unfil_percent[3]` %**).

@@

Using a cut-off value for the detection confidence of 0.1 (,i.e. discarding all detections with confidence below 0.1), this left us with a total of **`r dets_tb_cat_sum_fmt`** detections to sort through (see @fig-detections-histogram for an histogram of the detection scores, grouped by detection types). After applying this confidence threshold (decided on by the Wildlife Working Group) the total images to process for each category are: **`r dets_tb_cat_fmt[1]`** animals (**`r dets_tb_cat_percent[1]` %**); **`r dets_tb_cat_fmt[2]`** persons (**`r dets_tb_cat_percent[2]` %**); and **`r dets_tb_cat_fmt[3]`** vehicles (**`r dets_tb_cat_percent[3]` %**).

```{r echo=FALSE}
#| label: fig-detections-histogram
#| fig-cap: "Histogram of detection confidence scores produced for each bounding box predicted by Megadetector, showing a bimodal distribution. The colors represent the three categories of object identified by Megadetector: animals, persons and vehicles. Note that the hsitogram is truncated at our selected confidence threshold of 0.1."
#| fig.width: 8
#| fig.height: 4.5

@@

```{r echo=FALSE}
#| label: basics-stats-3

@@

@@

Out of the **`r dets_tb_cat_fmt[1]`** animal detections, **`r processed_fmt`** have been processed to date using our platform (or **`r processed_pct` %**) for a total of **`r processed_img_fmt`** images (i.e. **`r processed_img_pct` %** of the total amount of images, or **`r processed_img_up_pct` %** of the total amount of images uploaded to the platform).

As stated above, all images of a given burst are viewed if at least one image in the burst contains a detections with confidence above the threshold. This include potentially empty images. If we add the empty images viewed as part of a burst, it brings the total amount to **`r viewed_fmt`** images, (i.e. **`r viewed_pct` %** of the total amount of images, or **`r viewed_pct_up` %** of the total amount of images uploaded to the platform).

```{r echo=FALSE}
#| label: mans

@@

Out of the **`r processed_fmt`** detections, **`r dels_nb_fmt`** were false positives (or **`r dels_nb_pct` %**), where Megadetector identified something, but there was nothing. In addition, **`r non_mans_nb_fmt`** were true positives (or **`r non_mans_nb_pct` %**), to which **`r mans_nb_fmt`** false negatives were added (manually added bounding boxes, **`r mans_nb_pct` %**). Out of the **`r non_mans_nb_fmt`** true positives, **`r orig_tb_fmt[3]`** had to have their bounding box adjusted. The table below summarises information for the tagging progress.

```{r echo=FALSE}
#| label: tab-summary-table

@@

We are committed to producing a reproducible workflow. All code necessary to reproduce this report is available at an online [repository](https://github.com/StewartWILDlab/rofcamtrap). The repo makes use of [**mdtools**](https://github.com/StewartWILDlab/mdtools) module, a python-based command line tool for converting Megadetector outputs into LabelStudio inputs, and LabelStudio outputs into CSV tables.

@@

```{r echo=FALSE}
#| label: quality-control

# Only annotations for animals
# TODO investigate id of cut length 1
anns_wide_animals <- anns_wide_noempty |>
  dplyr::filter(species != 'Human') |> 
  dplyr::filter(species != 'Vehicle') |> 
  dplyr::mutate(id_cut_length = purrr::map_dbl(id , \(x) {
    x |> stringr::str_split("_") |> purrr::map_dbl(~length(.x))
  })) |> 
  dplyr::filter(id_cut_length > 1) |> 
  dplyr::mutate(camera_id = purrr::map_chr(id , \(x) {
    x |>
      stringr::str_split("_") |>
      purrr::map(~.x[[2]]) |>
      unlist()
  }))

@@

# TODO streamline exif so you dont have to join
anns_wide_animals_joined <- anns_wide_animals |>
  dplyr::left_join(exif, by = c("source_file", "image_id"))

stopifnot(dim(anns_wide_animals)[1] ==  dim(anns_wide_animals_joined)[1])
stopifnot((anns_wide_animals_joined$source_file |> is.na() |> sum()) == 0)

@@

# anns_wide_animals_joined_dets <- anns_wide_animals |> 
#   dplyr::right_join(dets_images, by = "image_id")

saveRDS(anns_wide_animals_joined_mut, "data/objects/annotations_animals.rds")
```

@@

```{r eval=FALSE, echo=FALSE}
# library(forcats)
library(ggplot2)
# anns_wide_animals_joined_mut |> 
#   ggplot(aes(x=date_time, y=species)) + geom_point()
# anns_wide_animals_joined_mut |> 
#   dplyr::filter(species == "Fisher") |> View()

@@

```{r eval=FALSE, echo=FALSE}
# for_vic <- anns_wide_animals_joined_mut |> 
#   dplyr::select(id,species,date_time, camera_id) |> 
#   dplyr::left_join(cams_sf |> dplyr::select(camera_id, plot_id) |> sf::st_drop_geometry(),
#                    by = "camera_id")
# tot_camera <- for_vic$camera_id |> unique() |> length()
# hit_rates <- for_vic |> dplyr::select(species, camera_id) |> 
#   dplyr::group_by(species) |> dplyr::summarise(n_cameras=length(unique(camera_id))) |> 
#   dplyr::ungroup() |> dplyr::mutate(rate = n_cameras/tot_camera*100) |> 
#   dplyr::arrange(dplyr::desc(rate))
# # det_history_two_weeks <- 
# 
# readr::write_csv(for_vic, "../data/derived_data/observations_by_camera_2023Aug15.csv")
# readr::write_csv(hit_rates, "../data/derived_data/hit_rates_by_species_2023Aug15.csv")
```

```{r echo=FALSE}
#| label: species

anns_wide_animals_table <- anns_wide_animals |>
  dplyr::select(camera_id, species) |> 
  dplyr::filter(species != "Unknown") |>
  table() 
anns_wide_animals_table <- anns_wide_animals_table > 0
anns_wide_animals_table <- anns_wide_animals_table + 0
anns_wide_animals_table <- anns_wide_animals_table |> 
  as.data.frame() |> 
  tibble::rownames_to_column("camera_id") |> 
  dplyr::arrange(camera_id)
anns_wide_animals_table_tb <- anns_wide_animals_table |> 
  dplyr::select(-camera_id)
rownames(anns_wide_animals_table_tb) <- anns_wide_animals_table$camera_id
```

@@

base <- readRDS(here::here(paper_dir, "base_map.rds"))
base +
  ggplot2::coord_equal(xlim = c(-88.5, -84),
                       ylim = c(50, 54)) +
  scatterpie::geom_scatterpie(data = cams_sf_with_props,
                              ggplot2::aes(x=X, y=Y, group=plot_id),
                              cols=names(cams_sf_with_props)[4:(4+tmp_col-1)],
                              sorted_by_radius = TRUE,
                              legend_name = "Species",
                              pie_scale = 1.5) +
  ggplot2::labs(x="Lat", y="Long") +
    ggplot2::geom_label(ggplot2::aes(x=x, y=y, label=ECOZONE_NA),inherit.aes = FALSE,
                      size = 2.5, data = regions_labels_bis)
```

@@

```{r, eval=FALSE}
#| label: heatmap
#| echo: false
#| eval: false

@@

# rast <- raster::rasterFromXYZ(preds)
# raster::plot(rast)
# rast[rast<0] <- NA
# raster::plot(rast)
```

```{r, eval=FALSE}
#| label: heatmap2
#| echo: false
#| eval: false
base_cut +
  ggplot2::geom_density2d_filled(
    data = cams_sf_sp_filt, ggplot2::aes(X, Y),
    binwidth = 0.05, alpha = 0.5)

@@

```{r eval=do_eval, echo=FALSE}
#| label: bubbleplots

@@

plot_list <- list()

for (sp in unique(cams_sf_sp_sum$species)) {

@@

  gg <- base_cut +
    ggplot2::geom_point(data = dat, col = "black",
                        ggplot2::aes(x = X, y = Y, size = count), 
                        inherit.aes = FALSE) +
    ggplot2::facet_wrap(~month) +
    ggplot2::labs(size = "Days \ndetected",
                  title = sp)
  if (sp %in% c("Caribou", "Sandhill crane", "Moose", "Wolf")) {
    plot_list[[sp]] <- gg
  }

  ggplot2::ggsave(gg, filename = paste0("figures/",sp,"_bplot.png"))
}

saveRDS(plot_list, here::here(paper_dir, "bbplot_list.rds"))
```

@@

```{r echo=FALSE}
plot_list <- readRDS(here::here(paper_dir, "bbplot_list.rds"))
plot_list["Caribou"]
```
```{r echo=FALSE}
plot_list["Sandhill crane"]
```
```{r echo=FALSE}
plot_list["Moose"]
```
```{r echo=FALSE}
plot_list["Wolf"]
```
